Symmetric Circuits for Rank Gates

### 2017-08-03 sections/symmetric-circuits.tex
Updated first four sections.

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background
Circuits are a well-studied model in complexity theory and, more recently, they
have been studied in the context of finite model theory. Circuits in the
literature () are usually defined as DAG's, with each vertex labelled by some
element of a basis. While not explicitly stated, a consequence of this
definition is that, when evaluating a gate, the Boolean function labelling this
gate should not depend on any ordering or labelling of its input. In other
words, it must be a symmetric function.

Motivated by the desire to study circuits with gates that compute rank, a
function which is not symmetric, we develop a general framework for studying
functions whose output depends on the structure of the input. This structure
also allows us to define and explore appropriate symmetries on the function.

With the necessary structure on the input of each Boolean function explicitly given, we then
develop a circuit model which incorporates this structure, and so allows for
circuits to be defined over Basis of functions that are not symmetric.

Finally, we prove that no family of Basis of symmetric functions can improve on
the expressive power of the circuit model of Anderson and Dawar \cite{AndersonD17}.
Together with the main theorem of this paper this implies that the more general
circuit models that are defined over basis of functions which are not symmetric
(with respect to the whole symmetric group) are strictly more powerful.

@@

\subsection{$\tau$-Symmetric Functions}
% note: include many-sorted signature in background
Let $\tau = (R, S, \nu)$ be a many-sorted, relational signature. Suppose some
ordering on the set of relations and sorts, so $R = \{R_1 \ldots, R_t\}$ and $S
= \{s_1, \ldots s_p\}$, and let $\nu(R_i) = (a^i_1, \ldots , a^i_{\arty(R_i)})$,
where $a^i_q \in S$ for all $q \in [\arty(R_i)]$. For each $s \in S$ let $A_{s}$
be a non-empty set, and let $A = A_{s_1} \times \ldots \times A_{s_p}$. Let
\[
  X^{\tau}_A:= A_{a^1_1} \times \ldots \times A_{a^1_{\arty(R_1)}} \times \ldots
  \times A_{a^t_1} \times A_{a^t_{\arty(R_t)}}.
\]

Let $G \leq \sym_{A_{s_1}} \times \ldots \times \sym_{A_{s_p}}$. There is a
natural action of $G$ on $X$. It is easy to see that if $\mathcal{C}_A$ is the
category with objects given by $\tau$-structures over $A$ and morphisms by the
action of $G$, and $\mathcal{C}_X$ is the category of subsets of $X^{\tau}_A$
with morphisms similarly given by the action of $G$, then these two categories
are equivalent. More informally, the subsets of $X^{\tau}_A$ encode the
$\tau$-structures over $A$. This relationship is bijective and the action of $G$
factors through this bijection.

\begin{remark}
  Is the categorical language unnecessary? It seemed the quickest way of saying
  what I wanted to say formally.
\end{remark}

\begin{remark}
  In the above section (and just below) we talk about the action of
  $\sym_{A_{s_1}} \times \ldots \times \sym_{A_{s_p}}$ on elements of $X^\tau_A$
  and structured functions. In the proof of the support theorem, when defining a
  definable matrix for converting circuits into formulas, and later on in this
  section, it is useful to speak more generally and instead look at bijections
  from $A_1, \ldots A_{s_p}$ to $B_1, \ldots, B_{s_p}$, thus allowing us to map
  between $X^\tau_A$ and $X^\tau_B$ and so between the associated structured
  functions. I've added this in later on in the section, but I still need to
  incorporate it into the original definition (it's a more general notion in a
  sense, and so I think it should be incorporated). I'll wait for feedback
  before doing this as I'm not sure this formulation will survive.
\end{remark}

Note that there is also a natural action on functions of the form $\vec{i}:
X^{\tau}_A \rightarrow \{0,1\}$ given by $\sigma \cdot \vec{i}(x) =
\vec{i}(\sigma x)$.

We now introduce a few useful notions of symmetry. We call a function of the
form $F:\{0,1\}^{X^{\tau}_A} \rightarrow \{0,1\}$ a \emph{$(\tau, A)$-structured
  function}, with $\tau$ called the \emph{signature of $F$} and $A$ \emph{the
  universe of $F$}. Call $F$ a \emph{structured function} if it is a $(\tau,
A)$-structured function for some appropriate $\tau$ and $A$.

We say a $(\tau, A)$-function is \emph{$(\tau, G)$-symmetric} if for all
$\vec{i} \in \{0,1\}^{X^\tau_A}$ and $\sigma \in G$, $f(\sigma \cdot \vec{i}) =
f(\vec{i})$. We say $F$ is \emph{$\tau$-symmetric} if it is $(\tau,
\sym_{A_{s_1}} \times \ldots \times \sym_{A_{s_p}})$-symmetric. Let $\tau = (S,
R, \nu)$, with $\vert S \vert \leq 2$, $R = {R'}$ and $\nu (R') = (s_1, s_2)$
(with $s_1$ and $s_2$ being the first and last elements of S in the ordering
respectively); if $F$ is $\tau$-symmetric we say $F$ is \emph{matrix-symmetric}.
If $\tau$ is single-sorted we call $F$ \emph{simple} and if it consists of a
single unary relation, we call $F$ \emph{unary}. We say $F$ is
\emph{graph-symmetric} if $F$ is simple and matrix-symmetric.

\begin{lem}
  Let $X^\tau_A$ be defined as above and let $F: \{0,1\}^{X^{\tau}_A}
  \rightarrow \{0,1\}$, then $F$ is a symmetric function (in the sense of
  Anderson and Dawar \cite{AndersonD17}) if and only if $F$ is a simple and
  unary.
\end{lem}

We briefly introduce notions of equivalence useful for comparing functions.
\begin{definition}
  Let $\tau = (R, S, \nu)$ be a many-sorted signature and let $A = A_1 \times
  \ldots A_s$ and $B = B_1 \times \ldots \times B_s$ be a product of non-empty
  sets. A \emph{sorted bijection} between $A$ and $B$ is a function bijections
  $f: A \rightarrow B$ such that $f (A_i) = B_i$ for all $i \in S$. There is an
  obvious action of $f$ that maps $X^\tau_A$ to $X^\tau_B$ and
  $\{0,1\}^{X^\tau_A}$ to $\{0,1\}^{X^\tau_B}$.

  Let $L: X^\tau_A \rightarrow H$ and $L': X^\tau_B \rightarrow H$ be two
  functions for some finite set $H$. We say that $L \sim L'$, or \emph{$L$ is
    equivalent to $L'$}, if there is a sorted bijection $f: A \rightarrow B$
  such that for all $x \in X^\tau_A$, $L(x) = L'(f (x))$.

  Let $F$ be a $(\tau, A)$-structured function and $G$ be a $(\tau,
  B)$-structured function. Then we say that $F \sim G$, or \emph{$F$ is
    equivalent to $G$}, if there is a sorted bijection $f: A rightarrow B$ such
  that for all $\vec{i} \in \{0,1\}^{X^\tau_A}$, $F (\vec{i}) = G(f(\vec{i}))$.
\end{definition}

\begin{remark}
  There is an obvious connection between $\tau$-symmetric functions and
  generalised quantifiers (or closed classes of structures). Should I include
  details about this connection? I also feel a lot of dirtiness might be
  avoidable if we instead recast everything in terms of generalised quantifiers.
  For one, generalised quantifiers give a natural way of defining a Boolean
  function for each input universe.
\end{remark}

@@

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis of structured functions and $\tau$ a set of
  relation symbols, we define a \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$
  computing a $q$-ary query $Q$ is a structure $\langle G, W, \Omega, \Sigma,
  \Lambda, L\rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $W \subseteq G \times G$, where $W$ is called the wires of the circuit.
    $(G,W)$ must be a directed acyclic graph. For $g \in G$ we $H_g := \{ h \in
    G: W(h,g)\}$ be the set of children of $g$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ which maps input gates to $\tau \uplus \{0,1\}$ and where $\vert
    \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1} (1) \vert \leq 1$ and the
    internal gates get mapped into $\mathbb{B}$. Gates mapped to $\tau$ are
    called relational gates and gates mapped to 1 or 0 are called constant
    gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_R)_{R \in
      \tau}$ where for each $R \in \tau$, $\Lambda_R$ maps each relational gate
    $g$ with $R = \Sigma (g)$ to the tuple $\Lambda_R (g) \in [n]^r$, where $r$
    is the arity of the symbol $R$. When no ambiguity arises we write $\Lambda
    (g)$ for $\Lambda_R (g)$.
  \item $L$ to each gate a labelling on its children. Let $\tau_g$ and $A_g$ be
    such that $\Sigma(g)$ is a $(\tau_g, A_g)$-structured function. Then $L(g)$
    is a surjection from $X^{\tau_g}_{A_g}$ to $H_g$. We call $L(g)$ the
    \emph{structure-labelling} of $g$.
  \end{itemize}
\end{definition}

Given some finite $\tau$-structure $\mathcal{A}$ of size $n$ and some bijection
$\gamma: U \rightarrow [n]$ and let $\gamma \mathcal{A}$ be the structure formed
by mapping the universe of $\mathcal{A}$ in accordance with $\gamma$. The
evaluation of some $(\mathbb{B}, \tau)$-circuit $C$ proceeds by recursively
evaluating gates. The evaluation of the gate $g$ using $\gamma$ with input
$\mathcal{A}$ is denoted by $C[\gamma \mathcal{A}](g)$. The evaluation of a gate
$g$ is given as follows.
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate labelled by a $(\tau_g, A_g)$-structured
  function $\Sigma(g)$, let $L^{\gamma}_g: X^{\tau_g}_{A_g} \rightarrow \{0,1\}$
  be defined by $L^{\gamma}_g(x) = C[\gamma \mathcal{A}](L(g)(x))$. Then $g$
  evaluates to true iff $\Sigma(g) (L^{\gamma}_g)$.
\end{enumerate}
$C$ defines the $q$-ary query $Q \subseteq \mathcal{A}^q$ where $a \in Q$ iff
$C[\gamma \mathcal{A}](\Omega (\gamma a)) = 1$.

@@

\begin{definition}[Invariant Circuit]
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary query. We
  say $C_n$ is \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of
  size $n$, $a \in \mathcal{A}^q$, and bijections $\gamma_1, \gamma_2: U
  \rightarrow [n]$ we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 a)) =
  C[\gamma_2 \mathcal{A}](\Omega (\gamma_2 a))$.
\end{definition}

This property ensures that a circuit (or a family of circuits) is invariant
under isomorphism. In other words, the circuit (or family of circuits) decides a
graph property. The following lemma allows us to recast this notion in terms of
the language developed in this paper.

\begin{definition}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit computing a Boolean query. We have
  that $C_n$ computes a $(\tau, [n])$-structured function $F$. We say $C_n$ is
  $\tau$-symmetric if $F$ is $\tau$-symmetric.
\end{definition}

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. $C_n$ is $\tau$-symmetric iff
  $C_n$ is invariant.
\end{lem}

The following definition gives us a way of saying that two structured functions
with the same signature are equivalent up to a choice of their universe.

\begin{definition}[Automorphism]
  let $C = \langle G, W, \Omega, \Sigma, \Lambda, L\rangle$ be a
  $(\mathbb{B},\tau)$-circuit computing at $q$-ary query on structures of size
  $n$. Let $\sigma \in \sym_n$ and $\pi: G \rightarrow G$ be a bijection such
  that
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item for all gates $g, h \in G$, $W(g,h)$ iff $W(\pi g, \pi h)$,
  \item for all output tuples $x \in [n]^q$, $\pi \Omega (x) = \Omega (\sigma
    x)$,
  \item for all gates $g \in G$, let $\Sigma (g) \sim \Sigma (\pi g)$,
  \item for each relational gate $g \in G$, $\sigma \Lambda (g) = \Lambda (\pi
    g)$, and
  \item for each internal gate $g$ if $\Sigma (g) \in \mathbb(B)$ then we have
    that $L(\pi g) \sim \pi \cdot L(g)$.
  \end{itemize}

  We call $\pi$ an \emph{automorphism} of $C$, and we say that $\sigma$
  \emph{induces the automorphism} $\pi$. The group of automorphisms of $C$ is
  called $\aut_n (C)$.
\end{definition}

@@

\begin{definition}[Rigidity]
  Let C be a $(\mathbb{B}, \tau)$-circuit, where $C = \langle G, W, \Omega,
  \Sigma, \Lambda, L\rangle$. Say that $C$ is rigid if there are no distinct
  internal gates $g, g' \in G$ such that $\Sigma(g) \sim \Sigma (g')$,
  $\Omega^{-1}(g) = \Omega^{-1}(g')$, $H_g = H_{g'}$ and $L(g) \sim L'(g)$.
\end{definition}

@@

We finally want to define a circuits that allow for gates labelled by
matrix-symmetric functions.
\begin{definition}
  Let $\mathbb{B}$ be a basis containing all simple unary structured functions
  that compute $\land$, $\lor$, $\neg$ and $\maj$ and such that all elements of
  this basis are either matrix-symmetric or simple unary. We call a
  $(\mathbb{B}, \tau)$-circuit a \emph{matrix-symmetric circuit}.
\end{definition}

We define a circuit that has gates that compute rank.
\begin{definition}
  Let the $(\mathbb{B}, \tau)$-circuit $C$ be a matrix-symmetric circuit. Let
  $\mathbb{B}_m$ be the subset of $\mathbb{B}$ consisting of all those functions
  that are matrix-symmetric but not symmetric. For any non-empty sets $A_1, A_2$
  and $r, p \in \mathbb{N}$, where $p$ prime define a $F_{p,r}: \{0,1\}^{A_1
    \times A_2} \rightarrow \{0,1\}$ such that $f_{p,r}(M) = 1$ iff the matrix
  $M: A_1 \times B_1 \rightarrow \{0,1\}$ has rank at least $r$ over the field
  $\mathbb{F}_p$. We call such a function a \emph{bounded-rank function}. If
  $\mathbb{B}_m$ is equal to the set of all such bounded-rank functions, then we
  call $C$ a \emph{symmetric circuit with rank gates}.
\end{definition}

@@

We are now ready to state what will be the main theorem of this paper.
\begin{thm}[Main Theorem]
  A graph property is decidable by a $P$-uniform family of symmetric circuits
  with rank gates if, and only if, it is definable by an FPR sentence.
\end{thm}

>>>>

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background
Circuits are a well-studied model in complexity theory, and more recently they
have been studied in the context of finite model theory. However, there are some
subtleties in the definition and evaluation of circuits which must be considered
carefully and have important implications for the notion of a symmetric circuit
studied in this paper. Let $C$ be a Boolean circuit $g$ be an internal gate
labelled by some function $f_g: \{0,1\}^{A} \rightarrow \{0,1\}$, for some index
set $A$ with $\vert A \vert = \vert H_g \vert$. Formally, evaluating $g$
requires a (usually bijective) indexing function $f_i: A \mapsto H_g$, which
organises the inputs of $g$ into an appropriate sequence. The evaluation of $g$
is then given by $C[\vec{x}](g) = f_g (a \mapsto C[\vec{x}](f_i(a)))$. As such
the evaluation of a gate in a circuit may depend on how the inputs to the gate
are indexed.

In most contexts in complexity theory, circuits may be taken to be implicitly
ordered objects and so this indexing function can be defined as taking the $i$th
element of $A$ to the $i$th element of $H_g$. However, in general this renders
the evaluation of a circuit for a given input a property both of the circuit and
an arbitrary order. In order to address this problem this problem many authors
only define circuits over Boolean bases of symmetric functions (e.g. the
standard basis or the majority basis). In this case the evaluation of a gate is
independent of the choice of indexing function.

However, in this paper we require gates labelled by Boolean rank functions, i.e.
a function that takes in a Boolean sequence indexed by `row' and `column' sets,
outputs 1 iff that matrix, when interpreted as having entries in $\mathbb{F}_p$,
has rank at least $r$. These rank functions clearly depend on how the input
sequence is labelled, and so are not symmetric. However, it is worth noting that
while such rank functions may be symmetric, they are symmetric in the weaker
sense of being constant under permutations on the rows and columns of the input
matrix.

Motivated by the desire to study circuits with gates that compute rank, We
develop a general framework for circuits that allows gates to be labelled by
Boolean functions of the form $F: \{0,1\}^{X} \rightarrow \{0, 1\}$, where the
labelling $X$ is the universe of a many-sorted structure over some vocabulary
$\tau$. The Moreover, we generalise the notion of a circuit for structures (see
Anderson and Dawar \cite{AndersonD17}), incorporate an appropriate labelling for
the inputs of each gate and allowing for a Boolean basis including non-symmetric
functions. Furthermore, the vocabulary $\tau$ will allow us to impose natural
symmetry conditions on $F$, and so allowing us to develop a useful notion of a
symmetric circuit.

We also prove that no family symmetric functions defined over a basis of
symmetric functions can improve on the expressive power of the circuit model of
Anderson and Dawar \cite{AndersonD17}. Together with the main theorem of this
paper this implies that these more general symmetric circuits are strictly more
powerful.

@@

\subsection{Structured Functions and Symmetry}
Let $\tau = (R, S, \nu)$ be a many-sorted, relational signature, with $R = \{R_1
\ldots, R_t\}$ and $S = [s]$ for some $s \in \nats$, and $\nu(R_i) = (a^i_1,
\ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
[r_i]$. For each $s \in S$ let $A_s$ be a non-empty set, and let $A := A_{1}
\sqcup \ldots \sqcup A_{s} = \{(a,s) : a \in A_s\}$. Let
\[
  \ind (A, \tau) := A_{a^1_1} \times \ldots \times A_{a^1_{r_1}} \times \ldots
  \times A_{a^t_1} \times A_{a^t_{r_t}},
\]. We call $\ind (A, \tau)$ the \emph{index defined by} $(A, \tau)$.

% It is worth noting that there is a natural bijection $P$ that maps
% $\tau$-structures with universe $A$ to subsets of $\Ind(A, \tau)$, taking each
% $\tau$-structure $\mathcal{A}$ to the set $\{(\vec{X}_1, \ldots , \vec{X}_t) :
% \exists i \in [t] \(\vec{X_i} \in R^{\mathcal{A}}_i \) \}$.

For the remainder of the section let $X := \ind (A, \tau)$. Let $G \leq
\sym_{X}$, there is a natural action of $G$ on $X$. There is also a natural
action of $\sym_{A_1} \times \ldots \times \sym_{A_s}$ on $X$ and a subgroup
$G'$ of $\sym_X$ corresponding to the same action. We identify these two groups.

Furthermore, we can defined an action of $G$ on functions of the form $f: X
\rightarrow H$, for any set $H$, given by $\sigma \cdot f(x) = f(\sigma \cdot
x)$ for all $x \in X$. If $H = \{0,1\}$ we say that $f$ is a \emph{binary
  sequence indexed by $X$}. If $H$ is surjective, we say that \emph{$f$ indexes
  $H$ by $X$} or \emph{$f$ indexes $H$ by $(A,\tau)$}.

We develop a notion of equivalence from this action. Say that two functions $f,
g : X \rightarrow H$ are \emph{$G$-equivalent} iff there exists $\sigma \in G$
such that $f(x) = f(\sigma \cdot x)$ for all $x \in X$. We say $f$ and $g$ are
\emph{sort-equivalent} if they are $\sym_{A_1} \times \ldots
\sym_{A_p}$-equivalent.

We call a function of the form $F:\{0,1\}^{X} \rightarrow \{0,1\}$ a
\emph{$(\tau, A)$-structured function}. We call $(\tau, A)$ the \emph{type} of
$F$, $\tau$ the \emph{signature of $F$} and $A$ \emph{the universe of $F$}. We
also use $\ind(F)$ to abbreviate $\ind (\type (F))$.Call $F$ a \emph{structured
  function} if it is a $(\tau, A)$-structured function for some appropriate
$\tau$ and $A$.

% It is easy to see that if $\mathcal{C}_A$ is the category with objects given
% by $\tau$-structures over $A$ and morphisms by the action of $G$, and
% $\mathcal{C}_X$ is the category of subsets of $X^{\tau}_A$ with morphisms
% similarly given by the action of $G$, then these two categories are
% equivalent. More informally, the subsets of $X^{\tau}_A$ encode the
% $\tau$-structures over $A$. This relationship is bijective and the action of
% $G$ factors through this bijection.

% \begin{remark}
%   Is the categorical language unnecessary? It seemed the quickest way of
%   saying what I wanted to say formally.
% \end{remark}

% \begin{remark}
%   In the above section (and just below) we talk about the action of
%   $\sym_{A_{s_1}} \times \ldots \times \sym_{A_{s_p}}$ on elements of
%   $X^\tau_A$ and structured functions. In the proof of the support theorem,
%   when defining a definable matrix for converting circuits into formulas, and
%   later on in this section, it is useful to speak more generally and instead
%   look at bijections from $A_1, \ldots A_{s_p}$ to $B_1, \ldots, B_{s_p}$,
%   thus allowing us to map between $X^\tau_A$ and $X^\tau_B$ and so between the
%   associated structured functions. I've added this in later on in the section,
%   but I still need to incorporate it into the original definition (it's a more
%   general notion in a sense, and so I think it should be incorporated). I'll
%   wait for feedback before doing this as I'm not sure this formulation will
%   survive.
% \end{remark}

These structured functions give us a general framework for indexing the input to
functions. Moreover, the structure on these index sets allows us to develop a
natural notion of symmetry particular to the indexing used by the function. Let
$F$ be a $(\tau, A)$-structured function. We say $F$ is \emph{$G$-invariant} if
for all $G$-equivalent $\vec{x},\vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) =
F(\vec{y})$. We say $F$ is \emph{sort-invariant} if it is $(\tau, \sym_{A_{1}}
\times \ldots \times \sym_{A_{s}})$-invariant. If $\tau = (\{R\}, [2], \nu)$,
with $\nu (R) = (1, 2)$ and $F$ is sort-invariant then we say $F$ is
\emph{matrix-invariant}.

% Let $F$ be a $(\tau, A)$-structured function with $\tau = (\{R\}, \{1\},
% \nu)$, where $R$ is a binary relation. Let $G \leq \sym_A$ such that for any
% $v, w \in A$ We say $F$ is \emph{graph-symmetric} if $F$ is $(\tau, A)$simple
% and matrix-symmetric.

% If $\tau$ is single-sorted we call $F$ \emph{simple} and if it consists of a
% single unary relation, we call $F$ \emph{unary-relational}.
We have developed this more general notion of symmetry for Boolean functions
that naturally takes into account the structure on the input sequence. The
following lemma shows that symmetry is just the requirement that the function be
invariant under permutations of its input, irrespective of structure.

\begin{lem}
  Let $F$ be a $(\tau, A)$-structured function. $F$ is a symmetric function if
  and only if $F$ is $\sym_A$-invariant.
\end{lem}

% We briefly introduce notions of equivalence useful for comparing functions.
% \begin{definition}
%   Let $\tau = (R, S, \nu)$ be a many-sorted signature and let $A = A_1 \times
%   \ldots A_s$ and $B = B_1 \times \ldots \times B_s$ be a product of non-empty
%   sets. A \emph{sorted bijection} between $A$ and $B$ is a function bijections
%   $f: A \rightarrow B$ such that $f (A_i) = B_i$ for all $i \in S$. There is
%   an obvious action of $f$ that maps $X^\tau_A$ to $X^\tau_B$ and
%   $\{0,1\}^{X^\tau_A}$ to $\{0,1\}^{X^\tau_B}$.

%   Let $L: X^\tau_A \rightarrow H$ and $L': X^\tau_B \rightarrow H$ be two
%   functions for some finite set $H$. We say that $L \sim L'$, or \emph{$L$ is
%   equivalent to $L'$}, if there is a sorted bijection $f: A \rightarrow B$
%   such that for all $x \in X^\tau_A$, $L(x) = L'(f (x))$.

%   Let $F$ be a $(\tau, A)$-structured function and $G$ be a $(\tau,
%   B)$-structured function. Then we say that $F \sim G$, or \emph{$F$ is
%   equivalent to $G$}, if there is a sorted bijection $f: A rightarrow B$ such
%   that for all $\vec{i} \in \{0,1\}^{X^\tau_A}$, $F (\vec{i}) =
%   G(f(\vec{i}))$.
% \end{definition}

% \begin{remark}
%   There is an obvious connection between $\tau$-symmetric functions and
%   generalised quantifiers (or closed classes of structures). Should I include
%   details about this connection? I also feel a lot of dirtiness might be
%   avoidable if we instead recast everything in terms of generalised
%   quantifiers. For one, generalised quantifiers give a natural way of defining
%   a Boolean function for each input universe.
% \end{remark}

@@

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis of structured functions and $\tau := (R, \{s_1,
  \ldots, s_p\}, \nu)$ be a many-sorted vocabulary and let $\vec{n} := (n_{s_1},
  \ldots , n_{s_p}) \in \mathbb{N}^{S}$, we define a \emph{$(\mathbb{B},
    \tau)$-circuit} $C_{\vec{n}}$ computing a $q$-ary query $Q$ of type
  $(s^Q_{1}, \ldots , s^Q_{q})$ is a structure $\langle G, W, \Omega, \Sigma,
  \Lambda, L\rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_{\vec{n}}$ and $\vert C_{\vec{n}}
    \vert := \vert G \vert$.
  \item $W \subseteq G \times G$, where $W$ is called the wires of the circuit.
    $(G,W)$ must be a directed acyclic graph. For $g \in G$ we $H_g := \{ h \in
    G: W(h,g)\}$ be the set of children of $g$.
  \item $\Omega$ is an injective function from $[n_{i_1}] \times \ldots \times
    [n_{i_q}]$ to $G$. The gates in the image of $\Omega$ are called the output
    gates. When $q = 0$, $\Omega$ is a constant function mapping to a single
    output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ which maps input gates to $\tau \uplus \{0,1\}$ and where $\vert
    \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1} (1) \vert \leq 1$ and the
    internal gates get mapped into $\mathbb{B}$. Gates mapped to $R$ are called
    relational gates and gates mapped to 1 or 0 are called constant gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R'})_{R' \in
      R}$ where for each $R' \in R$ with arity $r$ and type $(s^{R'}_{1},
    \ldots, s^{R'}_{r})$, $\Lambda_{R'}$ maps each relational gate $g$ with $R'
    = \Sigma (g)$ to the tuple $\Lambda_{R'} (g) \in [n_{s^{R'}_1}] \times
    \ldots \times [n_{S^{R'}_r}]$. When no ambiguity arises we write $\Lambda
    (g)$ for $\Lambda_{R'} (g)$.
  \item $L$ maps to each internal gate a labelling on its children. Let $g$ be
    an internal gate. We have that $L(g)$ is a surjection from $\ind (\Sigma (g))$
    to $H_g$. We call $L(g)$ the \emph{child-labelling} for $g$.
  \end{itemize}
\end{definition}

We write $C_{\vec{n}}$ to emphasise that the circuit accepts input structures
with sorted-size $\vec{n}$. If $\tau$ is single-sorted vocabulary then we write
$C_n$ to emphasise that $C_n$ accepts input structures of size $n$. For a gate
$g \in C_{\vec{n}}$ we abuse notation and define the \emph{type} of $g$, denoted
by $\type (g)$, to be type of $\Sigma(g)$, and the \emph{index} of $g$, denoted by
$\ind(g)$, to be the the index for $\Sigma(g)$.

Let $\tau = (\{s_1, \ldots, s_p\}, \{R_1, \ldots , R_t\}, \nu)$. Let
$\mathcal{A}$ be a many-sorted $\tau$-structure with sorts $U_1, \ldots, U_p$
and sort-size $\vec{n}$, and let $\gamma: U \rightarrow [n_{s_1}] \sqcup \ldots
\sqcup [n_{s_p}]$ be a sorted bijection. Let $\gamma \mathcal{A}$ be the
structure formed by mapping the universe of $\mathcal{A}$ in accordance with
$\gamma$. The evaluation of some $(\mathbb{B}, \tau)$-circuit $C$ computing a
q-ary query $Q$ of type $(s^Q_{1}, \ldots , s^Q_{q})$ proceeds by recursively
evaluating gates. The evaluation of the gate $g$ using $\gamma$ with input
$\mathcal{A}$ is denoted by $C[\gamma \mathcal{A}](g)$. The evaluation of a gate
$g$ is given as follows.
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate labelled by a $(\tau_g, A_g)$-structured
  function $\Sigma(g)$, let $L^{\gamma}_g: \ind(F) \rightarrow \{0,1\}$ be
  defined by $L^{\gamma}_g(x) = C[\gamma \mathcal{A}](L(g)(x))$. Then $g$
  evaluates to true iff $\Sigma(g) (L^{\gamma}_g)$.
\end{enumerate}
$C$ defines the $q$-ary query $Q \subseteq U_{s^Q_1} \times \ldots \times
U_{s^Q_q}$ where $\vec{a} \in Q$ iff $C[\gamma \mathcal{A}](\Omega (\gamma
\vec{a})) = 1$.

@@

\begin{definition}[Invariant Circuit]
  Let $C_{\vec{n}}$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary
  query of type $(s^Q_1, \ldots , s^Q_q)$. We say $C_{\vec{n}}$ is
  \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of size
  $\vec{n}$, $\vec{a} \in U_{i_1} \times \ldots \times U_{i_q}$, and sorted
  bijections $\gamma_1, \gamma_2: U \rightarrow [n_{s_p}] \sqcup\ldots \sqcup
  [n_{s_p}]$ we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 \vec{a})) =
  C[\gamma_2 \mathcal{A}](\Omega (\gamma_2 \vec{a}))$.
\end{definition}

This property ensures that the function computed by a circuit (or a family of
circuits) is invariant under isomorphisms on the input structure. In other
words, the circuit (or family of circuits) decides a property of the structure.
The following lemma allows us to recast this notion in terms of the language
developed in this paper.

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. We have that the function
  computed by $C_n$ is $\tau$-symmetric iff $C_n$ is invariant.
\end{lem}

For the sake of simplicity, from now on we assume the that all circuits take in
structures over a single-sorted vocabulary.

\begin{definition}[Automorphism]
  let $C = \langle G, W, \Omega, \Sigma, \Lambda, L\rangle$ be a
  $(\mathbb{B},\tau)$-circuit computing a $q$-ary query on structures of size
  $n$, and where $\mathbb{B}$ is a bases of sort-symmetric functions. Let
  $\sigma \in \sym_n$ and $\pi: G \rightarrow G$ be a bijection such that
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item for all gates $g, h \in G$, $W(g,h)$ iff $W(\pi g, \pi h)$,
  \item for all output tuples $x \in [n]^q$, $\pi \Omega (x) = \Omega (\sigma
    x)$,
  \item for all gates $g \in G$, let $\Sigma (g) = \Sigma (\pi g)$,
  \item for each relational gate $g \in G$, $\sigma \Lambda (g) = \Lambda (\pi
    g)$, and
  \item for each internal gate $g$ then we have that $L(\pi g)$ and $ \pi \cdot
    L(g)$ are sort-equivalent.
  \end{itemize}
  We call $\pi$ an \emph{automorphism} of $C$, and we say that $\sigma$
  \emph{induces the automorphism} $\pi$. The group of automorphisms of $C$ is
  called $\aut_n (C)$.
\end{definition}

@@

\begin{definition}[Rigidity]
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit, where $C_n = \langle G, W,
  \Omega, \Sigma, \Lambda, L\rangle$. Say that $C_n$ is \emph{rigid} if there
  are no distinct internal gates $g, g' \in G$ such that $\Sigma(g) = \Sigma
  (g')$, $\Omega^{-1}(g) = \Omega^{-1}(g')$, $H_g = H_{g'}$ and $L(g')$ and
  $L(g)$ are sort-equivalent.
\end{definition}

@@

We finally want to define a circuits that allow for gates labelled by
matrix-symmetric functions.

\begin{definition}
  Let $\mathbb{B}$ be a basis containing the single-sorted structured functions
  that compute $\land$, $\lor$, $\neg$ and $\maj$ and such that all elements of
  this basis are either matrix-symmetric or simple unary. We call a
  $(\mathbb{B}, \tau)$-circuit a \emph{symmetric matrix-circuit}.
\end{definition}

We define a circuit that has gates that compute rank.
\begin{definition}
  Let the $(\mathbb{B}, \tau)$-circuit $C$ be a matrix-symmetric circuit. Let
  $\mathbb{B}_m \subseteq \mathbb{B}$ consist of all matrix-symmetric functions
  that are not symmetric. For any non-empty sets $A_1, A_2$ and $r, p \in
  \mathbb{N}$, where $p$ prime define a $F_{p,r}: \{0,1\}^{A_1 \times A_2}
  \rightarrow \{0,1\}$ such that $f_{p,r}(M) = 1$ iff the matrix $M: A_1 \times
  B_1 \rightarrow \{0,1\}$ has rank at least $r$ over the field $\mathbb{F}_p$.
  We call such a function a \emph{bounded-rank function}. If $\mathbb{B}_m$ is
  equal to the set of all such bounded-rank functions, then we call $C$ a
  \emph{symmetric rank-circuit}.
\end{definition}

@@

We are now ready to state what will be the main theorem of this paper.
\begin{thm}[Main Theorem]
  A graph property is decidable by a $P$-uniform family of symmetric circuits
  with rank gates if, and only if, it is definable by an $\FPR$ sentence.
\end{thm}
